---
title: Processing ADCP Data with R
author: Masumbuko Semba
date: '2018-10-15'
slug: processing-adcp-data-with-r
categories:
  - Oceanography
tags:
  - ADCP
  - IIOE-2
  - Indian Ocean
  - Mafia Channel
bibliography: [argo.bib]
csl: apa.csl
link-citations: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(digits = 4)
```

### Introduction
The Second International Indian Ocean Expedition (IIOE-2) collected variaous oceanographic data within the coastal water of Tanzania. Among the oceanographic data collected using Agulhas II, is the Acoustic Doppler Current Profiles (ADCP). The measurement was carried out from 4^th^ to 9^th^ November 2017. In this post, I will illustrate how to process the ADCP data with R using `oce` package developed by @oce. First, we need to load other packages needed for this routine using `require()` function [@tidyverse; @ocedata; @lubridate; @leaflet; @sf]. 

```{r, echo=TRUE, warning=FALSE, message=FALSE, comment=""}
require(oce)
require(ocedata)
require(tidyverse)
require(lubridate)
require(leaflet)
require(sf)
```

### Ingestion of the data
Once the package needed for this routine are loaded, we can start processing the data. However, we need to import this dataset into the workspace using the `read.adp()` function from `oce` package
```{r, echo=FALSE}
## this chunk is the one loading the data, I muted it by pursing the argument echo = FALSE to prevent viewer to see the path of the file. the chunk below does not execute 
adp = read.adp("C:/Users/Semba/Documents/IIOE2/adcp/IIEO2/AGU028002_000000.LTA")
# summary(adp)
time = adp[["time"]]
```


```{r, echo=TRUE, eval=FALSE}
adp = read.adp("./AGU028002_000000.LTA")
# summary(adp)
time = adp[["time"]]
```

Figure \@ref(fig:fig1) show the velocity of zonal (north) and meridional (east) components from the raw data as captured by the ADCP instrument. These velocities were measured along the coastal of Tanzania water from `r time[1]` off Unguja Island close to Chwaka Bay and finish on `r time[length(time)]` in Mtwara.  

```{r fig1, fig.cap="Meridional and Zonal components velocity before averaging the profiles", , echo=TRUE}

plot(adp, which = c(1:2), 
     tformat = "%B %d, %Y:%H:%M", 
     missingColor = "grey", 
     ylim = c(25,1000),
     zlim = c(-4,6),
     useSmoothScatter = TRUE,
     ytype = "profile",
     titles = c("Northing Profiles","Northing Profiles" ),
     col = oceColorsViridis(125),
     marginsAsImage = TRUE, decimate = FALSE)
```

### Ensemble adp objects 
The grey-colored in figure \@ref(fig:fig1) are bad data mostly comes from single-ping velocity. Therefore, data are averaged to reducem the measurement uncertainity to acceptable levels [@oce]. Ensemble refers as a process of averaging ping profiles from ADCP data. This approach is important because it remove the uncertainty in velocity estimates from single pings, it also reduce the size of high frequency data, which reduce processing time. I used  `adpEnsembleAverage()` function from `oce` package to ensemble (average) the profiles as shown in the chunk below. 


```{r, echo=TRUE}
ensembled = adp%>%adpEnsembleAverage(n = 10, leftover = TRUE, na.rm = TRUE)

# summary(ensembled)

```

The ADCP object profiles were reduced by a factor of 10 and since they were average, the command `na.rm = TRUE` was parsed to ensure that profiles without values are also computed.
The origin adcp file had a total of `r adp@metadata$numberOfSamples` profiles, which were reduced to `r ensembled@metadata$numberOfSamples` with the `adpEnsembleAverage`. Figure \@ref(fig:fig2) show the velocity of zonal (north) and meridional (east) components of ensembled (average ping profiles) recorded by the ADCP instrument. 


```{r fig2, fig.cap="Meridional and Zonal components velocity after averaging the profiles", , echo=TRUE}
plot(ensembled, which = c(1:2), 
     tformat = "%B %d, %Y:%H:%M", 
     missingColor = "grey", 
     ylim = c(25,1000),
     zlim = c(-4,6),
     useSmoothScatter = TRUE,
     ytype = "profile",
     titles = c("Northing Profiles","Northing Profiles" ),
     col = oceColorsViridis(125),
     marginsAsImage = TRUE, decimate = FALSE)
```
### Depth cell mapping
The averaged profile (figure \@ref(fig:fig2)) were then aligned to the uniform depth---a process referred as `depth cell mapping`. Figure \@ref(fig:fig3) show the cells that aligned to the standard depth. This process is important because it ensure horizontal homogeneity of cells that are at the same depth, which compansate the pitch and roll of the instrument during ship cruising.  This only works for adcp data that are in beam coordinates. 

```{r, echo=TRUE}
ensembled.bin = binmapAdp(ensembled)

```


```{r fig3, fig.cap="Meridional and Zonal components velocity after averaging the profiles and align to standard depth", echo=TRUE}
## plot the ensembled
plot(ensembled.bin, which = c(1:2), 
     tformat = "%B %d, %Y:%H:%M", 
     missingColor = "grey", 
     ylim = c(25,1000),
     zlim = c(-4,6),
     useSmoothScatter = TRUE,
     ytype = "profile",
     titles = c("Northing Profiles","Northing Profiles" ),
     col = oceColorsViridis(125),
     marginsAsImage = TRUE, decimate = FALSE)
```





```{r, echo=FALSE, eval=FALSE}
#Get names of the Acoustic Doppler Beams
ensembled.bin%>%beamName()
```

### Extract the variables
Figure \@ref(fig:fig4) show the sampled location of ADCP profile pings.

```{r}

time = ensembled.bin[["time"]]
distance = ensembled.bin[["distance"]]
lon = ensembled.bin[["firstLongitude"]]
lat = ensembled.bin[["firstLatitude"]]

```

```{r, fig4, fig.cap="ADCP measurement along the coastal of Tanzania"}
position = data.frame(lon,lat, time)

leaflet(data = position) %>%
  addTiles() %>%
  addMarkers(lng = ~lon, lat = ~lat, popup = ~time)
```


### Subset ADCP
Since I am interested on the current between Lindi and Mafia Island (Figure \@ref(fig:fig5)), I ought to chop the pings that fall within the area. Unfortunately the function `subset()`, which I used, only capable of chopping adcp data using either time or distance. Based on figure \@ref(fig:fig4), only the pings profiles within the areas were selected by visual inspection of the points on the map and then use the date file to filter those pings with `subset()` function. Once the pings were selected, then `distance, longitude, latitude, v, u` were extracted. Using the `u` and `v` component, the `velocity` was computed with the equation \@ref(eq:eqn1)

$$
\begin{equation}
Velocity (ms^{-1})\:=\: \sqrt{(U^2+V^2)} (\#eq:eqn1)
\end{equation}
$$

```{r, echo=TRUE}
## creae a date limit
date = c("2017-11-07 13:14:52", "2017-11-07 23:14:51")%>%as_datetime()

## subset to the area based on date limit,
kilwa.adcp = ensembled.bin%>%subset(time >= date[1] & time <= date[2])%>%subset(distance < 700)

## extract variables from the subset
time = kilwa.adcp[["time"]]
distance = kilwa.adcp[["distance"]]
lon = kilwa.adcp[["firstLongitude"]]
lat = kilwa.adcp[["firstLatitude"]]
v = kilwa.adcp[["v"]][,,1]
u = kilwa.adcp[["v"]][,,2]

vel = sqrt(v^2 + u^2)

```

```{r fig5, echo=TRUE, fig.cap="A transect of ADCP between Lindi and Mafia Island"}
## make sf feature
kilwa.sf = data.frame(lon,lat,time)%>%st_as_sf(coords = c("lon", "lat"))%>%st_set_crs(4326)

## plot the location of ADCP
leaflet(data = kilwa.sf)%>%
  addTiles()%>%
  addMarkers(popup = ~time)
```
Figure \@ref(fig:fig6) show ocean current betwen Lindi and Mafia Island (figure \@ref(fig:fig5)) decrease with depth. The current at the surface is strong and decrease gradually with increase in depth. We also notice the current becomes more strong from Lindi toward Mafia, this can be contributed by the East African Coastal Current [@mcclanahan].


```{r fig6, fig.cap="Current Profile ", echo=TRUE}

## plot the velocity profiles as section
imagep(lat,distance, vel-6, filledContour = TRUE, 
       ylim = c(700, 30), zlim = c(0,1.0), zclip = TRUE, 
       col = oceColors9A(120), xlim = c(-9.5, -8.3), 
       xlab = "Latitude", ylab = "Depth (meters)", 
       zlab = "Current Velocity [m/s]", zlabPosition = "side")

```



```{r}
# create profiles
kilwa.profile = vel %>% 
   # transpose
  t() %>% 
  # convert to data frame
  as.data.frame() %>%
  # make it tibble
  as.tibble()

## change variable names
colnames(kilwa.profile) =  paste("profile", 1:12, sep = "")

## bind the disance to the profile
kilwa.profile = data.frame(distance, kilwa.profile)

```


```{r}
## make it long form from wide form
kilwa.profile.long = kilwa.profile %>%
  gather(key = "profile", value = "velocity", 2:13)

```

### Spurous measurements
We notice that the current measurements deviate from what has been reported previously. Figure \@ref(fig:fig7) show most profile along the transect from Lindi to Mafia have current speed above normal except profile measured at location with latitude `r lat[1]` and longitude `r lon[1]` (figure \@ref(fig:fig8)).  
```{r fig7, fig.cap="Current profiles at each ping"}
ggplot(data = kilwa.profile.long, aes(y = distance, x = velocity, col = profile))+
  geom_path()+
  scale_y_reverse()+
  scale_x_continuous(position = "top")+
  theme_bw()+
  theme(axis.text = element_text(colour = "black", size = 10),
        axis.title = element_text(colour = "black", size = 12),
        legend.position = "none")+
  labs(y = expression(~Distance~(m)),
       x = expression(~Current~Velocity~({ms^-1})))+
  facet_wrap(~profile, scales = "free_x")

```


```{r fig8, fig.cap="profile at station 1"}
ggplot(data = kilwa.profile.long %>%filter(profile %in% c("profile1")), aes(y = distance, x = velocity, col = profile))+
  geom_path()+
  scale_y_reverse()+
  scale_x_continuous(position = "top")+
  theme_bw()+
  theme(axis.text = element_text(colour = "black", size = 10),
        axis.title = element_text(colour = "black", size = 12),
        legend.position = "none")+
  labs(y = expression(~Distance~(m)),
       x = expression(~Current~Velocity~({ms^-1})))
```

### Absolute velocity
The ADCP measure a relative velocity to the earth, to obtain an absolute velocity, we need to substract the relative velocity with the ship velocity as illustrated in the chunk below

```{r, echo=TRUE}
##  calculate ship velocity
ship.vel = kilwa.adcp[["avgSpeed"]]
## obtain absolute velocity
vel.abs = abs(vel-ship.vel)

```
After correction, the absolute currrent velocity ranged from `r min(vel.abs)` ms^-1^  to `r max(vel.abs)` ms^-1^ with the median of `r median(vel.abs)` ms^-1^. We see current speed decrease with depth where the surface water has a relatively higher current speed than the deep water (figure \@ref(fig:fig9)). It also clear the surface current velocity increase from  Lindi to around latitude 9 ^o^S has relatively lower surface current and increases from 8 ^o^S toward Mafia (figure \@ref(fig:fig9)), which might be caused by East African Coastal Current.

```{r fig9, fig.cap="ADCP profile current Velocity between Lindi and Mtwara after correction of the ship speed", echo=TRUE}

## plot the velocity profiles as section

imagep(lat,distance, vel.abs, filledContour = TRUE, 
       ylim = c(30, 700), zlim = c(0,.6), zclip = TRUE, flipy = TRUE,
       col = oceColors9A(120), xlim = c(-9.5, -8.3), 
       xlab = "Latitude", ylab = "Depth (meters)", 
       zlab = "Current Velocity [m/s]", zlabPosition = "side",
       at = seq(0.04, 0.58, length.out = 6)%>%round(digits = 2), 
       drawPalette = TRUE, drawTriangles = TRUE)

```
### Cited Literature