---
title: Drawing Publication Quality maps with ggplot2 in R
author: Masumbuko Semba
date: '2018-10-20'
slug: genetic-connectivity-in-western-indian-ocean-region
categories: []
tags:
  - Pemba Channel
bibliography: [argo.bib]
csl: apa.csl
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = "")
```
### Moving from ARGIS to R
For years, ESRI\textsuperscript{\textregistered} ArcGIS was a core software for most data processing, spatial analysis and drawing of outputs as map. Honestly the elegant high-resolution figures and maps I produced did not come from ArcMap, but rather from Adobe Illustrator. I was doing all the layout of the maps in ArcMap and then export it as in vector format `Encapsulated PostScript`. I then imported the `.eps` file into Adobe illustrator for touching and polishing. Although ArcMap has many cartographic tools and abilities [@esri], but a lot of professional cartography is done in a vector graphics software such as Adobe Illustrator (and/or an image editing software such as Adobe Photoshop) [@illustrator] . 

The main reason for using Adobe Illustrator was that it is a more robust graphics software package and provides tools and effects not available in ArcGIS. These allow me to access and manipulate each and every object on the page as a graphic object, even the individual vertices of the text letters. Therefore, combining ArcMap and Adobe Illustrator provided me the best I wanted. But the labour invested in creating a single graph or map was painful. Another main reason for looking for an alternative was the license issue. Esri Arcmap and Adobe illustrator are proprietary software and hook your pocket for using them--- unable to purchase the license for using the software in developing countries. The other reason made me sick of the combinantion is when I was asked to make some edits on the graphics or maps created. I hated to go back and redo the same procedures to make the same map and accomodate the correction. My mind frequently flushed with an idea to make the procedure to make map and graphics once and then reproduce it. However, that was far from the reality because ArcGIS and Adobe Illustrator are two different platform. 

Switch to programming language and you are done! In  2015, I switched to R. One of the main reasons motivated me to turn to R is it's strong graphic power [@r] and reproducibility. In this post I will explain and show how to make elegant map with standard quality even for publishing.  We first need to load the packages we need for drawing maps. 

```{r}
require(tidyverse)
require(spData)
require(sf)

```

### Importing and Processing Spatial Data
A basemap was required that show the boundaries of the countries within the WIO region. The basemap that contain all african countries was used. Then load the basemap, for this post I used the basemap of Africa , which is the ESRI shapefile format. You can easily import shapefile into R workspace with the `read_sf()` function of `sf` package [@sf].
```{r eval=FALSE}
## read the shapefile
africa = read_sf("./africa/Spatial/AdmInfr/afcntry.shp")


```

Because I was working with the Western Indian Ocean region, I chopped the four countries of interest and dump the rest of the polygons in the layer with `filter()` function from `dplyr` package [@dplyr]. Trimming off unwanted polygons in a layer not only speed up drawing speed of the computer but also reduce processing time.
```{r eval=FALSE}
## read the shapefile
africa = read_sf("./africa/Spatial/AdmInfr/afcntry.shp")

## seclect the countries of Tanzania, Kenya and Mozambique nad Madagascar
wio = africa%>%
  select(-c(FIPS_CNTRY, REGIONA, EMPTY, EMPTY2))%>%
  filter(CNTRY_NAME == "Tanzania" | 
           CNTRY_NAME == "Kenya" | 
           CNTRY_NAME == "Mozambique" | 
           CNTRY_NAME == "Madagascar")


```

Once the African boundary shapefile was filtered too obtain the three boundaries of Kenya, Tanzania, Mozambique and Madagascar where the sampling was done. The geographical location of the sampling station was also ingested in R's worksapace. Since the file was in Excel spreashead, the `read_excel()` function from `readxl` package was used to import the file [@readxl]
 
```{r eval=FALSE}

# stations
point = readxl::read_excel("./Ogalo_mgeleka/stations.xlsx")

```


```{r include=FALSE, warning=FALSE, echo=FALSE}
#NOT these is hidden in the output, but it is the one that does the processing of the three chunk above which are simply for for display but never execute as I parsed the argument eval = FALSE in them

##read Africa continental shapefile that we will use throught the munuscript as basemap dataset for maps
africa = read_sf("E:/GIS/Tanzania spatial data Bank/EAF14 Tanzania spatial datasets/africa/Spatial/AdmInfr/afcntry.shp")

## clean the file
wio = africa%>%select(-c(FIPS_CNTRY, REGIONA, EMPTY, EMPTY2))%>%
  filter(CNTRY_NAME == "Tanzania" | CNTRY_NAME == "Kenya" | CNTRY_NAME == "Mozambique" | CNTRY_NAME == "Madagascar")

## station data
# stations
point = readxl::read_excel("E:/Data Manipulation/Ogalo_mgeleka/stations.xlsx")
```

Once the file was in the workspace, it was tied---cleaned and structured in a way make analysis and plotting easy. To make the data tidy and avoid clusters of points, only a single station was retained for areas with more than one station (Table \@ref(tab:tab1))
```{r tab1}

point.select = point  %>% 
  filter(Site %in% c("Inhaca South", "Inhaca North", "Nosy Beach East", " 
                     Pemba", "Mbweni")) 

point.select=point.select %>% 
  mutate(Site = replace(Site,Site == "Mbweni", "Zanzibar"),
         Site = replace(Site,Site == "Nosy Beach East", "Nosy Beach"))

point %>% kableExtra::kable("html", caption = "Geographical positions of the stations",
                col.names = c("Stations", "Longitude", "Latitude"), digits = 4, align = "c") %>%
  kableExtra::column_spec(column = 1:3, color = 1, width = "4cm") %>%
  kableExtra::add_header_above(c("", "Geographical Locations" =2), align = "c")
```

Then the tidy was converted from tabular formaat to simple feature---equvaivalent to popular spatial format shapefile. The transformation of tabular to shapefile was done using the combination of function from the `sf` package 
```{r}

point.sf = point %>% 
  st_as_sf(coords = c("Long", "Lat")) %>% 
  st_set_crs(4326)

```
 
 
### Mapping
Once the data was cleaned and converted to the right format, a `ggplot2` was used to create graphic.`geom_sf()` was used to draw maps the simple feature objects and limit the areas with `coord_sf()`. `geom_label_repel()` function from `ggrepel` package was used to label the sampling stations and the `scalebar()` function from `ggsn` package was used to draw scale on map. After the tweeking of different geom and them I ended up with map in figure \@ref(fig:map)
 
```{r map, fig.cap="map"}
wio.map = ggplot() + 
  geom_sf(data = spData::world, col = 1, fill = "ivory")+
  coord_sf(xlim = c(30,60), ylim = c(-30,5)) +
  geom_point(data = point.select, aes(x = Long, y = Lat), size = 2)+
  ggrepel::geom_text_repel(data = point.select, aes(x = Long, y = Lat, label = Site), nudge_y = 1.5, nudge_x = 5)+
  theme_bw() + 
  theme(panel.background = element_rect(fill = "lightblue"),
        axis.text = element_text(size = 11, colour = 1),
        panel.grid = element_line(colour = NA))+
  scale_x_continuous(breaks = c(35,55))+
  scale_y_continuous(breaks = c(-25, 2)) +
  ggsn::scalebar(location = "bottomleft", x.min = 35, x.max = 60,
                 y.min = -30, y.max = 5, dist = 600, dd2km = TRUE, 
                 model = "WGS84", st.dist = 0.02, st.size = 4)+
  labs(x = NULL, y = NULL)


wio.map

```

Though figure \@ref(fig:map) show the sampling areas within the Western Indian Ocean Region, its large scale masked subtle information of sampling station in these areas. For example, with figure \@ref(fig:map) we are able to show just a location in Zanzibar, but there were four sampling station in this area as revealed in zoomed portion of the island in figure \@ref(fig:unguja)

```{r unguja, fig.cap="Sampling stations at Unguja Island" }
zanzibar =ggplot()+ 
  geom_sf(data = wio, col = "black", fill = "ivory")+
  coord_sf(xlim = c(38.825, 39.6), ylim = c(-6.5, -5.63785)) +
  geom_point(data = point, 
  aes(x = Long, y = Lat), size = 2)+
  ggrepel::geom_text_repel(data = point  %>% 
  filter(Site %in% c("Nungwi", "Chwaka", "Fumba", "Mbweni")), 
  aes(x = Long, y = Lat, label = Site), point.padding = 0.2, nudge_y = 0.005, 
  nudge_x = -0.12, segment.colour = "red", direction = "y")+
  theme_bw() +
  theme(axis.text = element_text(size = 11, colour = 1),
        panel.background = element_rect(fill = "lightblue"),axis.title = element_blank(),
        panel.grid = element_line(colour = NA))+
  # scale_x_continuous(breaks = c(36.10, 39.55))+
  scale_y_continuous(breaks = c(-6.4, -5.7))+
  scale_x_continuous(breaks = c(38.9, 39.5))


zanzibar
```

Similar to Unguja, there were two stations at Madagascar, but in figure \@ref(fig:map) they clustered because of the small scale used and becomes difficult to present them on the map. However, when we use a large scale, we obtain a much clear graphic of the area (Figure \@ref(fig:nosy))
```{r nosy, fig.cap="Sampling stations at Nosy Madagascar" }

nosy = ggplot()+ 
  geom_sf(data = wio, col = "black", fill = "ivory")+
  coord_sf(xlim = c(48.38, 48.5), ylim = c(-13.58, -13.48)) +
  geom_point(data = point, 
  aes(x = Long, y = Lat), size = 4)+
  ggrepel::geom_text_repel(data = point  %>%
  filter(Site %in% c("Nosy Beach East", "Nosy Beach West")),
  aes(x = Long, y = Lat, label = Site), point.padding = 0.2, nudge_y = .0025, nudge_x = -0.03, segment.colour = "red")+
  theme_bw() +
  theme(axis.text = element_text(size = 11, colour = 1),
      panel.background = element_rect(fill = "lightblue"),axis.title = element_blank(),
      panel.grid = element_line(colour = NA))+
  scale_x_continuous(breaks = c(48.39, 48.49))+
  scale_y_continuous(breaks = c(-13.57, -13.49))

nosy


```


Further, there were two sampling points at Inhaca that are not visible in figure \@ref(fig:map).Zooming in at the location the distance between the points were clearly visible as shown in figure \@ref(fig:inhaca)

```{r inhaca, fig.cap="Sampling stations at Inhaca Mozambique"}

inhaca = ggplot()+ 
  geom_sf(data = wio, col = "black", fill = "ivory")+
  coord_sf(xlim = c(32.4, 33.17), ylim = c(-26.4,-25.62)) +
  geom_point(data = point,  aes(x = Long, y = Lat), size = 4)+
  geom_text(data = point, aes(x = Long, y = Lat, label = Site), nudge_x = -0.05, nudge_y = 0.05, size = 4.2)+
  theme_bw() +
  theme(axis.text = element_text(size = 11, colour = 1),
      panel.background = element_rect(fill = "lightblue"),axis.title = element_blank(),
      panel.grid = element_line(colour = NA)) +
  scale_x_continuous(breaks = c(32.5, 33.1))+
  scale_y_continuous(breaks = c(-26.3, -25.7))

inhaca


```
@cowplot developed a **cowplot** package which has tools to combine figures. For example the chunk below was used to combine figure \@ref(fig:cw1)

```{r cw1, fig.cap="The stations in A) Unguja-Tanzania and B) inhaca - Mozambique", fig.align="c"}

cowplot::plot_grid(zanzibar, inhaca, nrow = 1, labels = c("A", "B"), 
                   label_x = .16, label_y = .85, label_fontface = )
```

with **cowplot** package , it is possible to place several inset maps on the main map using the `ggdraw()` and `draw_plo()` functions. For example the chunk below produce the figure \@ref(fig:cw2).

```{r cw2, fig.cap="Map of area of interest drawn on on the WIO region area to show sampling stations", fig.width=8.5, fig.keep=TRUE}
cowplot::ggdraw()+ 
  cowplot::draw_plot(wio.map, x = 0, y = 0, width = 1, height = 1)+
  cowplot::draw_plot(nosy+theme(axis.text = element_blank(), axis.ticks = element_blank()), 
                     x = .51, y = 0.2,width = .6, height = .6,scale = .5)+ 
  cowplot::draw_plot(zanzibar+theme(axis.text = element_blank(), axis.ticks = element_blank()), 
                     x = .4, y = 0.49,width = .6, height = .6,scale = .6)+
  cowplot::draw_plot(inhaca+theme(axis.text = element_blank(), axis.ticks = element_blank()), 
                     x = .41, y = -0.05,width = .6, height = .6,scale = .6)
```

### Conclusion
We have seen the power of **R** and **ggplot2** to draw the publication quality graphics. With **cowplot** package, we can combine several graphics in a single layout. I have note done with Adobe illustrator---is still in high rank of software in my softwarebox. I still use this powerful software to edit and polish vector graphics generated from R like in figure\@ref(fig:cw3). The dotted line in figure\@ref(fig:cw3) are ocean current in the Western Indian Ocean region. The map was done in R, exported as `.eps` and ocean current---dotted line were drawn in **adobe illustrator** environement. 

```{r cw3, fig.cap="Ocean current in the WIO region", fig.height=7.5}
cowplot::ggdraw() +
  cowplot::draw_image("map.png" )
```


### Bibliography
