---
title: The Reasons for Creating my blog!
author: Masumbuko Semba
date: '2019-02-19'
slug: why-i-created-a-static-blog-from-scratch
categories:
  - Information Technology
tags:
  - Masumbuko Semba
  
bibliography: [blog.bib]
# csl: apa.csl
link-citations: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, comment = "")
```

According to [firstpage](https://www.firstpagemarketing.com/blog/the-importance-of-blogging/);

> ...blogging was initially used as a unique platform to share a person’s thoughts, feelings, opinions or experiences---an online journal or diary with a minimal following. 

Today, it has grown into a platform used by millions, everyone from individuals to businesses. The simplicity of a blog makes it easy for anyone to launch a blog and become a self-proclaimed “expert.” But how is it important in the world today, with unprecedented advent in technology and computing?

```{r, fig.align="center", echo=FALSE, eval=FALSE}
cowplot::ggdraw() +
  cowplot::draw_image("data_science_model.png")
```


After my first year chasing water---working on  the physical processes driving ocean current in the Indian Ocean, I ended up filling several notebooks and counterbooks with notes, script codes and graphics. During the course of a year I accumulate large volumes of diverse data on the hard drives of my computers. In addition to a variety of information relating to the study, such as the software, I also developed scripts for MATLAB and R programming that automate either the data scraping, processing, analysis and generating reports and published results. The rising flood of data and information acquired during the course of my study requires well-organized data file directories on the computer. 

Tracing these assets was hurdle and I had to spend several hours trying to figure out the location and working directories of each piece of work I had done previously. It was embarassing and thought wasnt health and effective way to store information!. The other subtle reason is that my brain tend to forget the code I write very soon and retrieving the things I have learned was hard. I had to look for a gatetway where I can post hard codes of the script I write and develop---a place that is easy to retrieve them. I got the idea to have my own blog---a platform that I will document most of the work I do and I can access---offline and online. It came out to be platform where I showcase my work, share content with other people dealing with physical oceanography and those like to script with R programming language. The constant demanding of writing a post every week has 

#### Developing the blog
I did not started the blog from scratch, I used the *hugo-template* and **blogdown** package developed by Xie and Thomas [-@blogdown] to create this blog. To make it less intimidating, I broke it down into pieces. The first stage I worked to simply develop a simple blog and deploy it. Then, once the blog was up and running, I had a glimpse of the blod and started working on the designing and appearance of the blog. Once the design was better and the quick loading of the pages, I started writing and pushing articles at least every week.

### hosting: Netlify
with so many tools available, its hard to know that right for the scope of the project. The important criteria for me were

+ Free hosting for a custom domain
+ Ease of use
+ Ability to serve dynamic
+ Built in HTTPS security

Netlify is a hosting and erverlesss backend service for static sites. It gave me everything I was looking for plus bonus features that I had not even thought of like

#### continuous deployment
Link your Github account and Netlify automatically built and deploy my blot. Every push of the updated blot in the Github is continuosuly deployed with Netlify
