---
title: Interactive map with leaflet
author: Masumbuko Semba
date: '2019-03-17'
slug: interactive-map-with-leaflet
categories:
  - Oceanography
tags:
  - CTD
  - Tanzania
  - Pemba Channel
  - Zanzibar
bibliography: [blog.bib]
# csl: apa.csl
link-citations: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, comment = "")
```

This post aim to train you how to make an interactive in R. We are going to focus on some poular packages for these task, many are part onf an **tidyverse** ecosystem [@tidyverse] with addition of **leaflet**[@leaflet] package. To work with these post you need to have version 3.4 and above of **R**[@r] installed in your machine. You also need to install the add-on packages **leaflet** and **tidyverse** if are not in your machine. We laod the packages we need for this task with `require()` function.


```{r}
require(tidyverse)
require(oce)
require(leaflet)
require(RColorBrewer)
```



```{r, echo=FALSE}

load("E:/Data Manipulation/ctd_bongo2018/ioe2")
pemba_zanzibar = section %>% subset( latitude >= -7) 
pemba_zanzibar_ctd = pemba_zanzibar[["station"]]
```

```{r, echo=FALSE}
ctd.tb = list()
for (i in 1:length(pemba_zanzibar_ctd)){

st = pemba_zanzibar_ctd[[i]]

ctd.tb[[i]] =  st@data %>%
  as_tibble()%>%
  select(pressure, temperature, fluorescence, oxygen) %>% 
  mutate(lon = st@metadata$longitude, lat = st@metadata$latitude, 
         date = st@metadata$startTime, station = i)
}

ctd.tb = ctd.tb %>% bind_rows()
```


```{r, echo=FALSE}
surface = ctd.tb %>% filter(pressure == 5)
```


Then we load the data from the local directory into R's workspace with the `read_csv()` function from **readr** package [@readr].
```{r, eval=FALSE}
surface = read_csv("./surface_ctd.csv")
```

 Table \@ref(tab:tab1) show the dataset contains six variable---temperature, fluorescence, oxygen all measured at water depth of 5 meter. The dataset also contains geographical (longitude and latitude) information, that we can used for making interactive map.
```{r tab1}
surface %>%
  select(1:6) %>%
  knitr::kable(format = "html", digits = 2, 
               col.names = c("Pressure", "Temperature", "Fluorescence", "Oxygen", "Longitude", "Latitude"),
               align = "c", 
               caption = "A table showing the variable in the dataset") %>%
  kableExtra::column_spec(column = 1:6, width = "4cm")
```

We map the location and show the variation of fluorence in each marker. We need to color the location markers with fluorescence values. To build a colour palette, we can use some utility functions in the **leaflet** package. The code below create a function that will generate a Yellow--Orange--Red palette from **RcolorBrewer** [@rcolorbrewer] package. The `domain` argument ensure that our colour scale will grade from the minimum to maximum fluorescence values.  

```{r}
oranges = colorNumeric(palette = "YlOrRd" ,
                       domain = range(surface$fluorescence))

```


We can now map the markers and specify the location markers coded with the `oranges` function of color pallete we just created
```{r}


leaflet(data = surface) %>%
  addProviderTiles("Esri.OceanBasemap") %>%
  addCircleMarkers(lng = ~lon, 
                   lat = ~lat, 
                   radius = 5, 
                   color = "grey80",  
                   weight = 1, 
                   fill = TRUE, 
                   fillOpacity = 0.7, 
                   fillColor = ~oranges(fluorescence)) 
```

We further add a legend to tell us what shade of color correspond to which value of fluorescence
```{r}


leaflet(data = surface) %>%
  addProviderTiles("Esri.OceanBasemap") %>%
  addCircleMarkers(lng = ~lon, 
                   lat = ~lat, 
                   radius = 8, 
                   color = "grey80",  
                   weight = 1, 
                   fill = TRUE, 
                   fillOpacity = 0.7, 
                   fillColor = ~oranges(fluorescence)) %>%
  addLegend(position = "topright", 
            pal = oranges, 
            values = surface$fluorescence, 
            title = "mg/L", 
            opacity = 1)
  
```

## References