---
title: Mapping with leaflets
author: Masumbuko Semba
date: '2019-10-02'
slug: mapping-with-leaflets
categories: []
tags: []
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Information of useful are found in this [link](https://cfss.uchicago.edu/notes/leaflet/) and read [this](http://www.econ.uiuc.edu/~lab/workshop/Spatial_in_R.html) for spatial modelling

```{r}
require(leaflet)
require(RColorBrewer)
require(tidyverse)

```

```{r}
wo = spData::world
```

```{r}
leaflet(data = wo) %>% 
  addPolygons(stroke = FALSE, fillOpacity = 0.5, smoothFactor = .5) %>% 
  addTiles()
```

The colorQuantile function of the leaflet package maps values of the data to colors following a palette. In this case I’ve specified a palette of Oranges and Reds, for more palettes you can access the help file for RColorBrewer: ?RColorBrewer.
```{r}

qpal<-colorQuantile("OrRd", wo$pop, n=5) 
```

Once the color is defined, we can use this color code to map the distribution of population in the world
```{r}


leaflet(data = wo)%>%
  setView(lng = 29.618994, lat = 0, zoom = 3) %>% 
  addPolygons(stroke = FALSE, fillOpacity = 0.7, smoothFactor = .9, color = ~qpal(pop)) %>% 
  addTiles()  %>%
  addLegend(pal = qpal,
            values = ~pop,
            opacity = 0.7,
            title = "world's Population",
            position = "bottomright")


```

