---
title: Creating wordcloud with wordcloud package in R
author: Masumbuko Semba
date: '2019-10-29'
slug: creating-wordcloud-in-r
categories:
  - R
tags:
  - wordcloud
  - R
bibliography: [blog.bib]
# csl: apa.csl
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


Google describe a wordclous as an image composed of words used in a particular text or subject, in which the size of each word indicates its frequency or importance. The words that appear more often are also the ones stand as bigger and bolder in wordcloud graphic.  You can create a word cloud, also referred as text cloud or tag cloud, which is a visual representation of text data.

The procedure of creating word clouds is very simple in R if you know the different steps to execute. @tm developed a text mining package (**tm**) to analyze texts and @wordcloud created a package to quickly visualize the keywords as a word cloud in R package.  In this article, I’ll demonstrate how analyse text and plot word clouds in R. 

## Why You Would Use word clouds ?
Wordcloud is used in various field including;

+ Researchers : for reporting qualitative data
+ Marketers : for highlighting the needs and pain points of customers
+ Educators : to support essential issues
+ Politicians and journalists
+ social media sites : to collect, analyze and share user sentiments


## packages
You need functions from various package. You must have these package installed in your machine. To install these package you use the `install.packages()` function as illustrated in the chunk below;

```{r, eval=FALSE}
install.packages("tm")  # for text mining
install.packages("SnowballC") # for text stemming
install.packages("wordcloud") # word-cloud generator 
install.packages("RColorBrewer") # color palettes

```

once you have installed these package, you can now load them to make the functions accessible for the task

```{r}
require("tm")
require("SnowballC")
require("wordcloud")
require("RColorBrewer")
require(tidyverse)
require(magrittr)

```

## Data
You need a text file with word that you need to make plot. For this post I process the *“I have a dream speech” from “Martin Luther King”* but you can use any text you want. Since the document is online, I used the `read_table()` function from **readr** package. 

```{r, echo=FALSE}

dream = read_table("./dream.txt")
```


```{r, eval=FALSE}
dream = read_table(file = "http://www.sthda.com/sthda/RDoc/example-files/martin-luther-king-i-have-a-dream-speech.txt", 
                   skip_empty_rows = TRUE,
                   progress = TRUE, 
                   comment = "" )
```


Then create a vector source of the text document we downloaded from the internet and transform it to corpus using `vectorSource()` and `Corpus()` function from the *tm* package [@tm]. 

```{r}
# Load the data as a corpus
# docs = Corpus(VectorSource(dream))

dream.corpus = dream  %>% 
  tm::VectorSource()%>% 
  tm::Corpus()
```

Once the corpus file is create, the `inspect()` function from **tm** package is used to show detailed information on a corpus document. 

```{r, eval=FALSE}
dream.corpus %>% inspect()
```

after converting to *corpus* dataset, we notice that some data cleaning is required. It appears from the earlier data inspection that there are many English words that we need to trim them off and retain only meaningful words. Remove words that include;

+ Numerical digits (0--9)
+ Stop words, which are common English words like *a* and *the*. If you do not remove from the dataset, they will dominate the word cloud over the meaningful words
+ remove punctuation and extra white space

The whole process is chained using the `%>%` operator from **magritr** package [@magrittr]
```{r}
dream.corpus.clean =  dream.corpus %>% 
  tm_map(FUN = content_transformer(tolower)) %>% # Convert the text to lower case
  tm_map(FUN = removeNumbers) %>% # Remove numbers
  tm_map(removeWords, stopwords("english")) %>% # Remove english common stopwords
  tm_map(removeWords, c("will", "let", "ring")) %>%   # Remove words 
  tm_map(removePunctuation) %>%   # Remove punctuations
  tm_map(stripWhitespace)   # Eliminate extra white spaces

```

```{r, eval=FALSE}
dream.corpus.clean %>% tm::inspect()
```

## Build a word frequency table
Once we have a clean, the function `TermDocumentMatrix()` from **tm** package [@tm] was used to make a matrix. Once the matrix is created is converted to data frame with `as.data.frame()` and pull the rownames into a data frame with the `rownames_to column()` function from **tibble** package [@tibble]. And finalize the wrangling with the `rename()` and `arrange()` function from **dplyr** package [@dplyr]. 

```{r}

dream.corpus.clean.tb=  dream.corpus.clean %>% 
  tm::TermDocumentMatrix() %>% 
  as.matrix() %>% as.data.frame() %>% 
  tibble::rownames_to_column() %>%
  dplyr::rename(word = 1, freq = 2) %>%
  dplyr::arrange(desc(freq))


```

Once the data is the right format, We plot the wordcloud with the wordcloud() function from **wordcloud** package [@wordcloud].

```{r, warning=FALSE, message=FALSE}


dream.corpus.clean.tb %$% 
  wordcloud::wordcloud(words = word, freq = freq,min.freq = 1, max.words = 200,
                       random.order = FALSE, rot.per = 0.35, colors = brewer.pal(8, "Dark2"))

```
Visualize the wordcloud graphic we get some key insights---The word freedom stands out compare to other words in the speech of Martin Luther King Junior on his speach of *I have a dream*

## References

