---
title: Wordclouds plotting with ggwordcloud package in R
author: Masumbuko Semba
date: '2019-11-05'
slug: wordclouds-plotting-with-ggwordcloud-package-in-r
categories: []
tags:
  - wordcloud
  - ggwordcloud
  - R
  - RStudio
  - Masumbuko Semba
  - Semba
  - Data science
  - programming
  - coding
bibliography: [blog.bib]
# csl: apa.csl
link-citations: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

[In the previous post](http://127.0.0.1:4321/10/29/2019/creating-wordcloud-in-r/), we saw how to make wordcloud graphic using the **wordcloud** package. In this post we extend the ability of R to make wordcloud graphic, but rather than using the base R function for plotting,I will show how to use **ggplot2** framework. 

@ggwordcloud developed a **gwordcloud** package, which extend the capability of **ggplot2** package [@ggplot] of making cloud text. With the `geom_text_wordcloud()` function, **gwordcloud** made making wordcloud plot easy with **ggplot2** package. The `geom_text_wordcloud()` function allows to grow the cloud according to a shape and stay within a mask. The size `aesthetic` is used either to control the font size or the printed area of the words. **ggwordcloud** also supports arbitrary text rotation. The faceting scheme of **ggplot2** can also be used. 

## packages
You need functions from various package. You must have these package installed in your machine. To install these package you use the `install.packages()` function as illustrated in the chunk below;

```{r, eval=FALSE}
install.packages("tm")  # for text mining
install.packages("SnowballC") # for text stemming
install.packages("wordcloud") # word-cloud generator 
install.packages("RColorBrewer") # color palettes
install.packages("ggwordcloud")
```

once you have installed these package, you can now load them to make the functions accessible for the task
```{r}
require("tm")
require("SnowballC")
require("wordcloud")
require("RColorBrewer")
require(tidyverse)
require(magrittr)
require(ggwordcloud)
```


## Data
You need a text file that store the word you need for visualization. For this post I'm going to process the *“I have a dream speech” from “Martin Luther King”* but you can use any text you want. Since the document is online, I used the `read_table()` function from **readr** package. 

```{r, echo=FALSE}

dream = read_table("./dream.txt")
```


```{r, eval=FALSE}
dream = read_table(file = "http://www.sthda.com/sthda/RDoc/example-files/martin-luther-king-i-have-a-dream-speech.txt", 
                   skip_empty_rows = TRUE,
                   progress = TRUE, 
                   comment = "" )
```


Then create a vector source of the text document we downloaded from the internet and transform it to a structure called *corpus* using `vectorSource()` and `Corpus()` function from the *tm* package [@tm]. 
```{r}
# Load the data as a corpus
# docs = Corpus(VectorSource(dream))

dream.corpus = dream  %>% 
  tm::VectorSource()%>% 
  tm::Corpus()
```

Once the corpus file is create, the `inspect()` function from **tm** package is used to show detailed information on a corpus document. 

```{r, eval=FALSE}
dream.corpus %>% inspect()
```

after converting to *corpus* dataset, we notice that some data cleaning is required. It appears from the earlier data inspection that there are many English words that we need to trim them off and retain only meaningful words. Remove words that include;

+ Numerical digits (0--9)
+ Stop words, which are common English words like *a* and *the*. If you do not remove from the dataset, they  dominate the word cloud over the meaningful words
+ remove punctuation and extra white space

The whole process is chained using the `%>%` operator from **magritr** package 
```{r}
dream.corpus.clean =  dream.corpus %>% 
  tm_map(FUN = content_transformer(tolower)) %>% # Convert the text to lower case
  tm_map(FUN = removeNumbers) %>% # Remove numbers
  tm_map(removeWords, stopwords("english")) %>% # Remove english common stopwords
  tm_map(removeWords, c("will", "let", "ring")) %>%   # Remove words 
  tm_map(removePunctuation) %>%   # Remove punctuations
  tm_map(stripWhitespace)   # Eliminate extra white spaces

```

```{r, eval=FALSE}
dream.corpus.clean %>% tm::inspect()
```

## Build a word frequency table
Once we have a clean, the function `TermDocumentMatrix()` from **tm** package [@tm] was used to make a matrix. Once the matrix is created is converted to data frame with `as.data.frame()` and pull the rownames into a data frame with the `rownames_to column()` function from **tibble** package [@tibble]. And finalize the wrangling with the `rename()` and `arrange()` function from **dplyr** package [@dplyr]. 

```{r}

dream.corpus.clean.tb=  dream.corpus.clean %>% 
  tm::TermDocumentMatrix() %>% 
  as.matrix() %>% as.data.frame() %>% 
  tibble::rownames_to_column() %>%
  dplyr::rename(word = 1, freq = 2) %>%
  dplyr::arrange(desc(freq))


```

```{r, warning=FALSE, message=FALSE}

ggplot(data = dream.corpus.clean.tb, 
       aes(label = word, size = freq, col = as.character(freq))) + 
  geom_text_wordcloud(rm_outside = TRUE, max_steps = 1,
                      grid_size = 1, eccentricity = .9)+
  scale_size_area(max_size = 14)+
  scale_color_brewer(palette = "Paired", direction = -1)+
  theme_void()
  
```

The words can be rotated by setting the angle aesthetic. For instance, one can use a rotation of 90 degrees for a random subset of 40 % of the words:
```{r}
 dream.corpus.clean.tb = dream.corpus.clean.tb %>%
   mutate(angle = 90 * sample(c(0, 1), n(), replace = TRUE, prob = c(60, 40)))

ggplot(data = dream.corpus.clean.tb, 
       aes(label = word, size = freq, angle = angle, col = as.character(freq))) + 
  geom_text_wordcloud(rm_outside = TRUE, max_steps = 1,
                      grid_size = 1, eccentricity = .9)+
  scale_size_area(max_size = 14)+
  scale_color_brewer(palette = "Paired", direction = -1)+
  theme_void()


```

## References

