---
title: Plotting Pie and Donut Chart with ggpubr pckage in R
author: Masumbuko Semba
date: '2020-04-06'
slug: plotting-pie-and-donut-chart-with-ggpubr-pckage-in-r
categories: []
tags:
  - pie chart
  - donut chart
  - ggpubr
  - ggplot2
  - analysis
bibliography: [blog.bib]
# csl: apa.csl
link-citations: yes
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = "")

```

Once in this blog I explained how to make [pie chart and donut plot with **ggplot2**](https://semba-blog.netlify.app/07/12/2019/pie-chart-and-donut-plot-with-ggplot2/). In this post, I will extend the topic by introducing **ggpubr** package, which has nifty function to facilitate a hustle free process of plotting publication quality figures. I will show you how you can use the function from this package to make a pie chart and donut chart with additional function to polish the charts from **ggplot2** [@ggplot]. First, we need to load the **tidyverse** package that will assist to process, manipulate and visualize the data [@tidyverse]. I prefer load packages using the `require()`, but you can use the `library` function, which does similar task. 

```{r}
require(tidyverse)

```

## Data
A pie chart need a series of data representing counts or proportions of different groups. Like the previous post, we download the daily logs of R version downloads with different operating systems. We use **cranlog** package [@cranlog] to obtain logs of R download for various operating system in the last thirty days from today.

```{r}
rdown = cranlogs::cran_downloads(packages = "R", 
                                 from = lubridate::today()-30, 
                                 to = lubridate::today()) %>%
  mutate(os = replace(os, os=="src", "Linux"),
         os = replace(os, os=="osx", "Mac"),
         os = replace(os, os=="win", "Windows"))

rdown %>% sample_n(size = 10)
```
To understand the R version download, we first ask the question, Are R downloads differs over weekdays and operating system? Here we want to understand whether there are days of week with more downloads than others and if there are which operating system has more downloads. We can answer this question by simply extract days of week using the `wday` function from **lubridate** package [@lubridate] and then group the compute the frequency of downloads based on the week day and operating system. Data without type of operating system were omitted from the dataset.The chunk below illustrate the code of lines used to prepare the data to answer the question asked above.

```{r }

weeday.logs = rdown %>% 
  mutate(day = lubridate::wday(date, abbr = TRUE, label = TRUE)) %>% 
  group_by(os,day) %>% 
  count() %>% 
  ungroup() %>% 
  filter(os != "NA") 

```

Figure \@ref(fig:fig1) shows the frequency of R downloads over week days at three different operating systems. We notice that while Windows and Linux download patterns are very close over the week days, the downloads from Mac operating system is higher than those from Windows and Linux operating system over all the days of the week. This suggest that Mac users downloaded more R from CRAN than any other operating system over the last 30 days. 

```{r fig1,fig.cap="Week day R downloads across multiple operating systems", fig.width=8, fig.height=4}
weeday.logs%>% 
  ggplot()+
  geom_col(aes(x = day, y = n, fill = os), position = position_dodge(.9))+
  ggsci::scale_fill_jama(name = "Operating \nSystem")+
  ggpubr::theme_pubclean()+
  theme(axis.title.x = element_blank(), legend.position = "right", 
        legend.key = element_blank())+
  coord_cartesian(expand = FALSE)+
  scale_y_continuous(breaks = seq(50,260,50))+
  scale_x_discrete(limits = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))+
  labs(y = "Number of downloads")
```

Then we might be interested the percentage of dominance of Mac in downloading R as compared to Linux and Windows Operating Systems. To be able to address this issue, we need first to compute the frequency of downloads for each operating system. The combination of `group_by` and `count` functions from **dplyr** [@dplyr] package helps us getting the frequencies of operating systems downloads. Once we computed the frequency, we can transform them into fraction and change into percentage. To be able to label, we stitch the percentage values and the operating system into a variable called label using the `paste` function. The chunk below highlight the key steps. 

```{r}
os = rdown %>% group_by(os) %>% 
  count() %>% 
  ungroup() %>% 
  filter(os != "NA") %>% 
  mutate(percentage = n/sum(n)*100) %>%
  mutate_if(is.numeric, round, digits = 0) %>%
  mutate(label = paste(os,"\n","(", percentage, "% ",")", sep = ""))

```

## Pie and donut chart
Although **ggplot** does a decent way to plot both the pie and donut plot, Alboukadel Kassambara -@ggpubr developed a **ggpubr** package, which extend some functions of **ggplot2**. Some of the function of **ggpubr** are `ggpie` and `ggdonutchart`, which have some arguments to pass on and generate a pie and donut plot directly. These function works well with other **ggplot2** functions and other tidyverse functions. The code below illustrate how to make figure \@ref(fig:fig2) using the `ggpie` function. 

```{r fig2, fig.cap="Pie chart from **ggpubr** package"}
ggpubr::ggpie(data = os, x = "n", label = "label", color = "white",
                     lab.font = "white",fill = "os",lab.pos = "in", 
                     orientation = "horizontal", lab.adjust = 100)+
  ggsci::scale_fill_jama()+
  theme(legend.position = "none")
```

Also the code below show how to use `ggdonutchart` function to generate a donut plot shown in figure \@ref(fig:fig3).

```{r fig3, fig.cap="Pie chart from **ggpubr** package"}
 ggpubr::ggdonutchart(data = os, x = "n", label = "label", color = "white",
                     lab.font = "white",fill = "os",lab.pos = "in", 
                     orientation = "horizontal")+
  ggsci::scale_fill_jama()+
  theme(legend.position = "none")


```

## Remarks
Although the post use **cranlog** package to download the logs of R across multiple operating system, but the this post mainly focus on using functions from **ggpubr** package to plot pie and donut chart. Until now I hope you have seen how easy to make pie and donut chart in R by combining **ggplot2** and **ggpubr** functions. I hope you wont be troubled anymore with how you should arrange the label in either pie or donut chart when using the **ggpubr** package, as it does well labeling the pieces. Figure \@ref(fig:fig4) and \@ref(fig:fig5) are bonus plots in this post made with **plotly** package. I have included the code for generating these plots in this post. Check for coming posts that explain making interactive plots with **plotly** package [@plotly]

```{r fig4, fig.cap="Pie chart from plotly package"}
plotly::plot_ly(data = os, 
                labels = ~label, 
                values = ~n, 
                type = 'pie',
                textinfo='label+percent',
                showlegend = FALSE) %>% 
plotly::layout(title = 'Percentage of R downloads by Operating System',
         xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
         yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE))
```



```{r fig5, fig.cap="Donut chart from plotly package"}
plotly::plot_ly(data = os, 
                labels = ~label, 
                values = ~n, 
                type = 'pie', hole = 0.6,
                textinfo='label+percent',
                showlegend = FALSE) %>% 
plotly::layout(title = 'Percentage of R downloads by Operating System',
         xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
         yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE))
```


## References

