---
title: 'Plots in Interactive Maps with R '
author: Masumbuko Semba
date: '2020-06-13'
slug: plots-in-interactive-maps-with-r
categories:
  - R
  - Oceanography
tags:
  - Masumbuko Semba
  - Semba
  - mapping
  - R
bibliography: [blog.bib]
# csl: apa.csl
link-citations: yes
---

## Introduction

Often times when w are working with data, there always a geospatial component to the data---the locations. R has several packages that support interactivity of features, including maps. @minicharts developed **leaflet.minicharts** package that enhance interactivity of maps by superimpose on the basemaps with interactive plots. In this post I illustrate tools *leaflet.minicharts** use case in which you have a set of species and you want an interactive map.

leaflet.minicharts is an R package that provides two functions to add and update small charts on an interactive maps created with the package leaflet. These charts can be used to represent as many variables as desired associated to geographical points. Currently, three types of chart are supported: barcharts (the default), pie charts and polar area charts.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = "")

```


```{r}

require(leaflet.minicharts)
require(leaflet)
require(sf)
require(tidyverse)
```


We first create a basemap that will be used in all the following examples

```{r}
tilesURL = "http://server.arcgisonline.com/ArcGIS/rest/services/Canvas/World_Light_Gray_Base/MapServer/tile/{z}/{y}/{x}"

basemap = leaflet(width = "100%", height = "800px") %>%
  addTiles(tilesURL)

basemap
```

## Census Data

Is the proportional of married people similar across all regions in Tanzania? To answer this question let us focus on the census data of 2002. We first import a simple feature into our session using `st_read` function from **sf** package [@sf].

```{r, echo=FALSE}
region = st_read("e:/GIS/Tanzania spatial data Bank/NBS correct/Tanzania_Region_EA_2002_region/Tanzania_Region_EA_2002_region.shp", quiet = TRUE)
```


```{r, eval=FALSE}

region = st_read("NBS correct/Tanzania_Region_EA_2002_region/Tanzania_Region_EA_2002_region.shp")

```

```{r}
region %>% glimpse()
```

A glimpse reveal a subtle internal structure of the dataset. It's a simple feature with polygons representing the geographical boundary of 26 regions in Tanzania. Each region contains demographic information collected in 2002 Housing and Population Census. We are only interested with number of people by gender and marital status, which we select them using `select` function from **dplyr** package [@dplyr]. Before the variable were selected, colnames were cleaned using `clean_names()` function from **janitor** package [@janitor].  

```{r}
region = region %>% 
  janitor::clean_names() %>%  
  select(regname,male,female,single, maried) %>% 
  st_transform(4326)
```

Since we need the centroid points of the region, `st_point_on_surface` function from **sf** package was used to convert polygon to point feature [@sf]. Then longitude and latitude of centroid points were extracted using `st_coordinates` function and combined with the attributes information to produce a tibble. Once the tibble was obtained, a total population in each region was computed by summing the number of male and female in each region. 

```{r}
region.tb = region %>% 
  st_point_on_surface() %>% 
  st_coordinates() %>% 
  as_tibble() %>% 
  rename(lon = 1, lat = 2) %>% 
  bind_cols(region %>% st_drop_geometry()) %>%
  mutate(total = male+female)
```

We now add to the base map a pie chart for each region that represents the share of married people. We also change the width of the pie charts so their area is proportional to the total population in the region.

```{r}

colors = c("#104E8B", "#FF00FF") #<- c("#4fc13c", "#cccccc")
colors2 <- c("#3093e5", "#fcba50")

basemap %>%
  addMinicharts(lng = region.tb$lon, 
                lat = region.tb$lat, 
                type = "pie", 
                chartdata = region.tb[, c("single", "maried")], 
                colorPalette = colors2, 
                width = 60 * sqrt(region.tb$total) / sqrt(max(region.tb$total)), 
                transitionTime = 0)
```

Now let’s represent the marital status using bar charts.

```{r}
basemap %>%
  addMinicharts(lng = region.tb$lon, 
                lat = region.tb$lat, 
                type = "bar", 
                chartdata = region.tb[, c("single", "maried")], 
                colorPalette = colors2, width = 25, height = 65)
```


## Representing a single variable

**leaflet.minicharts** has been designed to represent multiple variables at once, but you still may want to use it to represent a single variable. In the next example, we represent the total population of each region in 2002. When data passed to `addMinicharts` contains a single column, it automatically represents it with circle which area is proportional to the corresponding value. In the example we also use the parameter `showLabels` to display rounded values of the variable inside the circles.

```{r}

basemap %>%
  addMinicharts(lng = region.tb$lon, 
                lat = region.tb$lat, 
                # type = "bar", 
                chartdata = region.tb$total, 
                showLabels = TRUE, 
                width = 55)
```

## References

